---
title: "Building an R Package"
subtitle: "Introduction"
author: |
  | Steve Pederson
  | Bioinformatics Hub
  | University of Adelaide
date: "5 April 2016"
output: ioslides_presentation
---

## Introduction

Today's outline:

- Data Structures in R
- Writing Functions in R
- Documenting Functions in a Package
- Exporting To GitHub

## Required Packages

```{r, eval = FALSE}
install.packages(c("roxygen2", "devtools", "knitr", 
                   "dplyr", "magrittr"))
```

- Much of today's material is from *Advanced R* by Hadley Wickham, Chief Scientist at RStudio
    - The online version  is freely available at [http://adv-r.had.co.nz/](http://adv-r.had.co.nz/)

- Knowledge of R Studio is assumed  
- We will use an `R project` for today's session  

## R Project Setup

1. `File > New Project ...` 
2. `New Directory > Empty Project`  
3. Choose somewhere and call the project something relevant:
    + e.g. PackageBuilding  

## A file for taking notes

Open a new R Markdown File for taking notes  

1. `File > New File > R Markdown...`
    - Enter `Notes from Package Building` as the title
    - Leave it as an HTML output
    - Call it "notes.Rmd"
2. Delete everything except the YAML header (between the dashed lines)

## Using R Markdown

- We'll use this for taking notes  
- Plain text comes out as plain text
- To create a code chunk enter `Ctrl + Alt + I`
- Code in a chunk can be run using `Ctrl + Alt + C`

- **We can describe something, then put the code in a chunk**
    - Hitting the `Knit HTML` button at the top compiles a nice HTML document
    - Can also select `.pdf` & `.docx` for output
    - The Markdown Quick Reference Guide is opened using the `?` symbol at the top of the Source Editor

## Data in R

- Most commonly we use `data.frame` objects

- Underlying these are the key building blocks

- **Everything in R is a vector.**

- 4 `atomic` types:
    - `logical`, `integer`, `double`, `character`
    
- **Every component of an atomic vector is of the same data type!**

## Vectors

A simple vector could be
```{r}
1:10
```

All vectors have a `length()` attribute, but **no other** dimensions.


## Working with Vectors

We can *combine* two vectors in *R*, using the function `c()`

```{r}
c(1, 2)
```

- The numbers `1` & `2` were both vectors of `length` 1

- We have combined two vectors of `length` 1, to make a vector of `length` 2


## The 4 Atomic Vector Types

1. **logical**

```{r}
logi_vec <- c(TRUE, TRUE, FALSE)
print(logi_vec)
```

*Can only hold the values* `TRUE` *or* `FALSE`

## The 4 Atomic Vector Types

1. logical
2. **integer**

```{r}
int_vec <- 1:5
print(int_vec)
```

*Useful for counts, ranks or indexing positions, e.g. column 3, nucleotide 254731*

## The 4 Atomic Vector Types

1. logical
2. integer
3. **double**

```{r}
dbl_vec <- c(0.618, 1.414, 2)
print(dbl_vec)
```

*Often (& __lazily__) referred to as numeric*

## The 4 Atomic Vector Types

1. integer
2. logical
3. double
4. **character**

```{r, include=TRUE, results='markup'}
char_vec <- c("blue", "red", "green")
print(char_vec)
```

*Holds text strings*

## The 4 Atomic Vector Types

These are the basic building blocks for all *R* objects

1. logical
2. integer
3. double
4. character

There are two more rare types we'll ignore: `complex` & `raw`

Many other vector types built on these `atomic` vectors

## Properties of a vector

What defining properties might a vector have?

There are four...

## Properties of a vector

What defining properties might a **vector** have?

1. The actual *values*
2. *Length*, accessed by the function `length()`
3. The *type*, accessed by the function `typeof()`
    + Similar but preferable to `class()`
4. Any optional & additional *attributes* such as names, etc.; `attributes()`

## Properties of a vector

Let's try them on our vectors

```{r, results='hide'}
typeof(char_vec)
length(int_vec)
attributes(logi_vec)
class(dbl_vec)
typeof(dbl_vec)
```

*Were you surprised by the results?*

## Working with Vectors

**What would happen if we combined two vectors of different types?**

Let's try & see what happens:

```{r, results='hide'}
new_vec <- c(logi_vec, int_vec)
print(new_vec)
typeof(new_vec)
```

*What happened to the `logical` values?*

## Working with Vectors

**What would happen if we combined two vectors of different types?**

Let's try & see what happens:

```{r, results='hide'}
new_vec <- c(logi_vec, int_vec)
print(new_vec)
typeof(new_vec)
```

*What happened to the* `logical` *values?*

- `R` will **coerce** them into a common type.

## Coersion

### Discussion Questions
1. What other types could logical vectors be coerced into?
2. Can character vectors be coerced into numeric vectors?
3. Can numeric vectors be coerced into logical vectors?

Try using the functions: 
`as.integer()`, `as.logical()`, `as.double()` & `as.character()`

## Coersion

```{r}
simp_vec <- c(742, "Evergreen", "Terrace")
simp_vec <- as.numeric(simp_vec)
```

## Subsetting Vectors

The elements of a vector can be called using `[]`

```{r}
y <- c("A", "B", "C", "D", "E")
y[2]
y[c(1, 3)]
```

## Subsetting Vectors

Double brackets (`[[]]`) can be used to return single elements only

```{r, eval =1}
y[[2]]
```

If you tried `y[[c(1,3)]]` you would receive an error message

## Subsetting Vectors

If a vector has name attributes, we can call values by name

```{r}
euro[1:5]
names(euro)[1:2]
euro[c("ATS", "BEF")]
```

## Subsetting Vectors

*Try repeating the* call-by-name *approach using double brackets*

```{r, results='hide'}
euro["ATS"]
euro[["ATS"]]
```

**What was the difference in the output?**

1. Using `[]` returned the vector with the identical structure
2. Using `[[]]` removed the `attributes` & just gave the value

## Vector Operations

`R` Functions are designed to work on vectors

```{r, results='hide'}
dbl_vec - 1
dbl_vec > 1
dbl_vec^2
mean(dbl_vec)
sd(dbl_vec)
sqrt(int_vec)
```

This is one of the real strengths of `R`

## Vector Operations

We can combine the above logical test and subsetting

```{r}
dbl_vec[dbl_vec > 1]
```

This gives a further useful way of subsetting any vector

## Vector Operations

An additional logical test is `%in%`

- This can be read as *is in*

```{r}
dbl_vec %in% int_vec
```

Returns a `TRUE/FALSE` value based on whether each value in `dbl_vec` **is in** `int_vec`

NB: For this comparison `int_vec` would have been coerced silently to a `double` vector

## Matrices

Vectors are strictly one dimensional and have a `length` attribute.

A `matrix` is the two dimensional equivalent

```{r}
int_mat <- matrix(1:6, ncol=2)
print(int_mat)
```

## Matrices

- Matrices can only hold **one** type of value 
    + i.e. logical, integer, double, character
- Each row or column is essentially a vector
- Have additional attributes such as `dim()`, `nrow()` `ncol()`
- Optionally can have `rownames()` & `colnames()`

*What would you expect `length()` to return for a matrix?*

## Matrices

- We use square brackets to extract individual entries by row & column
- The form is `x[row, col]` where the comma divides the rows from columns

```{r, results='hide'}
int_mat[2, 2]
int_mat[1,]
```

*How would we just get the first column?*

## Matrices

**NB: Forgetting the comma when subsetting will treat the matrix as a single vector broken into columns**

```{r}
int_mat
int_mat[5]
```

*What would happen if we'd added a comma before or after* `5`*?*

## Arrays

Arrays extend matrices to 3 or more dimensions

Beyond the scope of today, but we just have more commas in the square brackets, e.g.
```{r}
dim(iris3)
```

Vectors, Matrices & Arrays are the basic *homogeneous* data types of `R`

## Heterogeneous Data types

Summary of main data types in *R*

Dimension   | Homogeneous | Heterogeneous
------------|-------------|--------------
1d          | `vector`    | `list`       
2d          | `matrix`    | `data.frame`
3d+         | `array`     |