---
title: "Building an R Package"
subtitle: "Lists and Data Frames"
author: |
  | Steve Pederson
  | Bioinformatics Hub
  | University of Adelaide
date: "5 April 2016"
output: ioslides_presentation
---

## Lists

A `list` is a *heterogeneous* vector.

- Each component is an `R` object type
- Can be a `vector`, or `matrix`
- Could be another `list`
- Any other `R` object type we haven't seen yet

```{r, echo=FALSE, include=TRUE, results='hide'}
logi_vec <- c(TRUE, TRUE, FALSE)
int_vec <- 1:5
dbl_vec <- c(0.618, 1.414, 2)
```

## Lists

Many `R` functions provide output as a list

```{r, results='hide'}
testResults <- t.test(dbl_vec)
typeof(testResults)
```

## Lists

Explore the various attributes of the object `testResults`

- `attributes()`; `length()`; `names()`

We can also call the individual components of `testResults` using the `$` symbol

- `testResults$statistic`; `testResults$conf.int`; `testResults$method`

Note that each component is quite different to the others.

*Use the function* `typeof()` *on some of the components above*

## Subsetting Lists

A `list` is a `vector` so we can also subset using the `[]` method

```{r, results='hide'}
testResults[1]
```

Double brackets can retrieve a single element

```{r, results='hide'}
testResults[[1]]
```

## Subsetting Lists

Using single square brackets returns a `list` with the structure intact

```{r}
typeof(testResults[1])
```

Using double square brackets returns the actual component as a distinct `R` object

```{r}
typeof(testResults[[1]])
```

*When would we use either method?*

## Lists

- Note also the **Environment** Tab in the top right of RStudio
- Click the arrow next to `testResults` to expand the entry
- This is the output of `str(testResults)`

## Data Frames

Finally!

- These are the *most common type of data* you will see
- Each column is a vector
- Columns can be different types of vectors

## Data Frames

- Analogous to matrices, but are specifically for *heterogeneous* data
- Have many of the same attributes as matrices
    + `dim()`, `nrow()`, `ncol()`, `rownames()`, `colnames()`
- `colnames()` & `rownames()` are **NOT** optional & are assigned by default

## Data Frames

Here's an example `data.frame` from the package *datasets*

Use `?ToothGrowth` to find out what's in the object

```{r}
head(ToothGrowth)
```

## Data Frames

Try these commands

```{r, results='hide'}
colnames(ToothGrowth)
dim(ToothGrowth)
nrow(ToothGrowth)
```

## Data Frames

Individual entries can also be extracted using the square brackets, as for matrices

```{r}
ToothGrowth[1:2, 1]
```

We can also refer to columns by name

```{r}
ToothGrowth[1:2, "len"]
```


## Data Frames

The concept of columns being distinct vectors is quite important & useful

- We can call each column vector of a `data.frame` using the `$` operator

```{r} 
ToothGrowth$len[1:2]
```

**This does NOT work for rows!!!**

## Data Frames

- `R` is *column major* by default (as is `FORTRAN` & Mat lab)
- Many other languages are *row major*, e.g. C/C++, Python
- `R` was designed for statistical analysis, but has developed capabilities **far beyond this**


## Data Frames & Lists

**Data frames are actually special cases of lists**

- Each column of a `data.frame` is a component of a `list`
- The components must all be vectors of the same length
- Lists and Data Frames behave identically, particularly when subsetting

## Data Frames & Lists

Forgetting the comma, now gives a completely different result to a matrix!

```{r, eval=FALSE}
ToothGrowth[1]
```

*Was that what you expected?*

Try using the double bracket method


## Creating `R` Objects

### Vectors

We've created many vectors already

- The important step is the `<-` symbol (or `->`)

### How do we assign names?

```{r}
named_vec <- c(a=1, b=2, c=3)
```

OR we can name an existing vector

```{r}
named_vec <- int_vec
names(named_vec) <- c("a", "b", "c", "d", "e")
```

## Creating `R` Objects

*Can we remove names?*

The `NULL`, or empty, vector in `R` is created using `c()`

```{r, results='hide'}
null_vec <- c()
length(null_vec)
```

We can use this to remove names

```{r}
names(named_vec) <- c()
```

*Don't forget to put the names back...*

## Creating `R` Objects

### Data Frames
This is exactly the same as creating lists, but

- The `names` attribute will also be the `colnames()`

```{r}
my_df <- data.frame(doubles = dbl_vec, logical = logi_vec)
names(my_df) == colnames(my_df)
```

## Creating `R` Objects

### Data Frames
*What happens if we try to add components that aren't the same length?*

`my_df` `<- data.frame(integers = int_vec,` 
                    `doubles = dbl_vec, logical = logi_vec)`
                    
